---
title: Linkage Disequilibrium Test for Microhaplotype Data
author: Alfred Hubbard
output:
  html_notebook: default
  pdf_document: default
  github_document: default
---

```{r setup, echo = FALSE, message = FALSE, warning = FALSE}
# Load required libraries ----------------------------------------------
# These will be referenced without the `package::` construct, and thus 
# are loaded second to avoid masking
library(knitr)
library(magrittr)
library(tidyverse)

# Don't include code in PDF output
opts_chunk$set(echo = ! is_latex_output(), out.width = "100%")
# Set default ggplot2 theme
theme_set(theme_bw())

# Hardcode parameters. These are not passed as shell arguments because 
# hardcoded values facilitate development and this document is not 
# meant to be usable with other projects.
inp <- list(ld_res = "../../results/ld_res.rds")
```

To evaluate linkage disequilibrium (LD) between pairs of loci, the $\bar{r}_{d}$ 
measure was used in favor of the more common index of association. The index of 
association is a ratio of the variance observed between individuals at each 
locus and the variance expected under conditions of linkage equilibrium. This 
measure increases with the number of loci, so $\bar{r}_{d}$ was developed as an
alternative that approximates the index of association while avoiding this
limitation.

LD was estimated using the populations defined by the MalariaGEN authors. A 
correction for clonal samples was performed prior to making the LD calculation, 
because having multiple copies of the same multi-locus genotype can bias the 
calculation. Significance was estimated using permutations. 

The distribution of *p*-values obtained between all sample pairs is shown below.


```{r warning = FALSE}
# Read and tidy LD results ---------------------------------------------
pairwise_ld_res <- read_rds(inp$ld_res) %>%
  as_tibble(rownames = "locus_pair") %>%
  mutate(across(-locus_pair, as.double)) %>%
  separate(locus_pair, c("locus_a", "locus_b"), sep = ":")

# Visualize distribution of p-values -----------------------------------
pairwise_ld_res %>%
  ggplot(mapping = aes(x = p.rD)) +
  geom_histogram(bins = 20) +
  labs(x = "p-value", y = "No. of Locus Pairs")
```

A uniform distribution is expected if the null hypothesis of no linkage is true. 
This histogram shows a distinct right skew, suggesting there may be linkage in 
our dataset.

However, many hypothesis tests were performed to obtain these results, so before 
concluding that any pairs are in significant LD, a correction for multiple 
testing should be performed. This was done with the Bonferroni method. This 
approach is sometimes considered too conservative, so a generous *p*-value of 
0.05 was used to help compensate for that possibility. The Bonferroni-corrected 
significance threshold for this case is:

```{r}
# Compute and apply Bonferroni correction for multiple testing ---------
bf_thres <- 0.05 / nrow(pairwise_ld_res)
bf_thres
pairwise_ld_res_signif <- pairwise_ld_res %>%
  filter(p.rD < bf_thres)
```

There are `r nrow(pairwise_ld_res_signif)` loci pairs that have significant LD 
at this threshold. Therefore, no filtering is necessary to address LD.
