---
title: Quantification of Microhaplotypes Identified by AmpSeq
author: Alfred Hubbard
output:
  html_notebook: default
  pdf_document: default
  github_document: default
---

```{r setup, echo = FALSE, message = FALSE, warning = FALSE}
# Load required libraries ----------------------------------------------
# These will be referenced without the `package::` construct, and thus 
# are loaded second to avoid masking
library(knitr)
library(magrittr)
library(readxl)
library(tidyverse)

# Don't include code in PDF output
opts_chunk$set(echo = ! is_latex_output())
# Set default ggplot2 theme
theme_set(theme_bw())

source("../shared_functions.R")

# Hardcode parameters. These are not passed as shell arguments because 
# hardcoded values facilitate development and this document is not 
# meant to be usable with other projects.
inp <- list(
  reads_summary_uci1223 = "../../results/AmpSeq/UCI1223/run_dada2/reads_summary.txt", 
  cigar_table_uci1223 = "../../results/AmpSeq/UCI1223/asv2cigar/seqtab_cigar.tsv", 
  locus_metadata = "../../results/locus_metadata.tsv", 
  rep_metadata_uci1223 = "../../resources/MH_template__LB_EHS_dz07242024_to_liz.xlsx", 
  reads_summary_serialdil2 = "../../results/AmpSeq/serialdil2/run_dada2/reads_summary.txt", 
  cigar_table_serialdil2 = "../../results/AmpSeq/serialdil2/asv2cigar/seqtab_cigar.tsv"
)
out <- list(
  rl_read_counts = "../../results/AmpSeq/serialdil2/rl_read_counts.csv", 
  rl_pct_good_amp_parasitemia = 
    "../../results/AmpSeq/serialdil2/rl_pct_good_amp_parasitemia.csv"
)
```

# Final Data Quantity

## UCI 12/23

### Reads

#### By Replicate

```{r fig.width = 15, fig.height = 17, message = FALSE, out.width = "100%"}
# Read in final, filtered data -----------------------------------------
variant_read_counts_uci1223 <- read_tsv(
    inp$cigar_table_uci1223, 
    col_types = cols(.default = col_integer(), sample = col_character()), 
    progress = FALSE
  ) %>%
  pivot_longer(-sample, names_to = "locus_cigar", values_to = "n_read") %>%
  separate_wider_delim(locus_cigar, ",", names = c("locus_pos", "cigar")) %>%
  # Remove strand information
  mutate(locus_pos = str_remove(locus_pos, "\\(.\\)")) %>%
  rename(sample_id = sample)

# Check for implicit missing data --------------------------------------
n_loci_by_rep_uci1223 <- variant_read_counts_uci1223 %>%
  group_by(sample_id) %>%
  summarize(n_loci = n_distinct(locus_pos), .groups = "drop")
reps_wo_all_loci_uci1223 <- n_loci_by_rep_uci1223 %>%
  filter(n_loci < n_distinct(variant_read_counts_uci1223$locus_pos))
if (nrow(reps_wo_all_loci_uci1223) > 0) {
  stop(
    "Not all replicates have entries for all loci in AmpSeq output", 
    call. = FALSE
  )
}

# Summarize read counts ------------------------------------------------
rl_read_counts_uci1223 <- variant_read_counts_uci1223 %>%
  # Calculate total reads for each replicate/locus
  group_by(sample_id, locus_pos) %>%
  summarize(n_read = sum(n_read), .groups = "drop")

# Read and join locus names --------------------------------------------
locus_metadata <- read_tsv(
    inp$locus_metadata, 
    col_types = cols(
      .default = col_character(), 
      start_pos = col_integer(), 
      end_pos = col_integer()
    ), 
    progress = FALSE
  ) %>%
  select(locus, chrom, start_pos, end_pos) %>%
  unite(start_end, start_pos, end_pos, sep = "-", remove = FALSE) %>%
  unite(locus_pos, chrom, start_end, sep = ":")
rl_read_counts_uci1223 <- rl_read_counts_uci1223 %>%
  left_join(locus_metadata, by = "locus_pos") %>%
  select(-locus_pos, -start_pos, -end_pos)

# Add missing loci that received no amplification ----------------------
rl_read_counts_uci1223 <- rl_read_counts_uci1223 %>%
  mutate(locus = factor(locus, levels = unique(locus_metadata$locus))) %>%
  complete(sample_id, locus, fill = list(n_read = 0)) %>%
  mutate(locus = as.character(locus))

# Read replicate metadata ----------------------------------------------
rep_metadata_uci1223 <- read_xlsx(
    inp$rep_metadata_uci1223, 
    sheet = "pv242"
  ) %>%
  select(
    SampleID, 
    `library name`, 
    MH_DNA_Well, 
    DARC_Phenotype, 
    Site, 
    Method, 
    `Ct...21`, 
    `Ct...23`
  ) %>%
  rename(
    sample_id = SampleID, 
    lib_name = `library name`, 
    specimen_id = MH_DNA_Well, 
    ct1 = `Ct...21`, 
    ct2 = `Ct...23`
  ) %>%
  pivot_longer(c(ct1, ct2), names_to = "ct_rep", values_to = "ct") %>%
  mutate(ct = na_if(ct, "No Ct")) %>%
  mutate(ct = as.numeric(ct)) %>%
  group_by(sample_id, lib_name, specimen_id, DARC_Phenotype, Site, Method) %>%
  summarize(ct = mean(ct), .groups = "drop")
# Make sure each sample ID has one and only one Ct value
sample_ct_values_uci1223 <- rep_metadata_uci1223 %>%
  distinct(specimen_id, ct) %>%
  group_by(specimen_id) %>%
  mutate(n_ct = n()) %>%
  ungroup()
if (max(sample_ct_values_uci1223$n_ct) > 1) {
  stop("Some sample IDs have more than one Ct value", call. = FALSE)
}
# Join metadata
rl_read_counts_uci1223 <- rl_read_counts_uci1223 %>%
  left_join(rep_metadata_uci1223, by = "sample_id")

# Plot reads by replicate and locus ------------------------------------
for (l in unique(rl_read_counts_uci1223$lib_name)) {
  lib_reads <- rl_read_counts_uci1223 %>%
    filter(lib_name == l)
  samples_reps <- lib_reads %>%
    distinct(specimen_id, sample_id) %>%
    arrange(specimen_id)
  lib_reads %>%
    arrange(specimen_id) %>%
    # Make sample_id a factor with ordering corresponding to samples_reps
    mutate(sample_id = factor(sample_id, samples_reps$sample_id, ordered = TRUE)) %>%
    read_count_heatmap(y_lbl = "Replicate", title_lbl = l) %>%
    print()
}
```

These heatmaps show the number of reads in the final dataset for each
locus-replicate combination. The plots have been grouped according to library. 
LB2 contained samples from Duffy positives, sequenced in duplicate, while LB5 
and LB7 contained samples from Duffy negatives that were sequenced 5x.

Sequencing yield was reasonably consistent across replicates, though there were 
a few exceptions to this. Sequencing yield was a lot better with samples from 
Duffy positive individuals, which is presumably related to parasitemia.

#### Sample Means

```{r fig.width = 15, fig.height = 11, out.width = "100%"}
# Compute mean read counts for each sample/locus -----------------------
sl_mean_read_counts_uci1223 <- rl_read_counts_uci1223 %>%
  filter(DARC_Phenotype != "negative") %>%
  select(-DARC_Phenotype) %>%
  group_by(specimen_id, locus, lib_name) %>%
  summarize(n_read = mean(n_read), .groups = "drop")

# Plot reads by sample and locus ---------------------------------------
sl_mean_read_counts_uci1223 %>%
  rename(sample_id = specimen_id) %>%
  read_count_heatmap() %>%
  print()
```

These heatmaps show the mean number of reads for each sample-locus combination.
Samples from Duffy negative individuals have been removed.

## Serial Dilution Run 2

### Reads

#### By Replicate

```{r fig.width = 15, fig.height = 17, message = FALSE, out.width = "100%"}
# Read in final, filtered data -----------------------------------------
variant_read_counts_serialdil2 <- read_tsv(
    inp$cigar_table_serialdil2, 
    col_types = cols(.default = col_integer(), sample = col_character()), 
    progress = FALSE
  ) %>%
  pivot_longer(-sample, names_to = "locus_cigar", values_to = "n_read") %>%
  separate_wider_delim(locus_cigar, ",", names = c("locus_pos", "cigar")) %>%
  # Remove strand information
  mutate(locus_pos = str_remove(locus_pos, "\\(.\\)")) %>%
  rename(sample_id = sample) %>%
  separate_wider_delim(
    sample_id, 
    "-", 
    names = c("well", "specimen_id", "parasitemia", "protocol"), 
    cols_remove = FALSE
  ) %>%
  mutate(parasitemia = as.integer(parasitemia))

# Check for implicit missing data --------------------------------------
n_loci_by_rep_serialdil2 <- variant_read_counts_serialdil2 %>%
  group_by(sample_id) %>%
  summarize(n_loci = n_distinct(locus_pos), .groups = "drop")
reps_wo_all_loci_serialdil2 <- n_loci_by_rep_serialdil2 %>%
  filter(n_loci < n_distinct(variant_read_counts_serialdil2$locus_pos))
if (nrow(reps_wo_all_loci_serialdil2) > 0) {
  stop(
    "Not all replicates have entries for all loci in AmpSeq output", 
    call. = FALSE
  )
}

# Summarize read counts ------------------------------------------------
rl_read_counts_serialdil2 <- variant_read_counts_serialdil2 %>%
  # Calculate total reads for each replicate/locus
  group_by(sample_id, well, specimen_id, parasitemia, protocol, locus_pos) %>%
  summarize(n_read = sum(n_read), .groups = "drop")

# Join locus names -----------------------------------------------------
rl_read_counts_serialdil2 <- rl_read_counts_serialdil2 %>%
  left_join(locus_metadata, by = "locus_pos") %>%
  select(-locus_pos)

# Add missing loci that received no amplification ----------------------
rl_read_counts_serialdil2 <- rl_read_counts_serialdil2 %>%
  mutate(locus = factor(locus, levels = unique(locus_metadata$locus))) %>%
  complete(
    nesting(well, specimen_id, parasitemia, protocol, sample_id), 
    locus, 
    fill = list(n_read = 0)
  ) %>%
  mutate(locus = as.character(locus))

# Print read count summary statistics ----------------------------------
summary(rl_read_counts_serialdil2$n_read)

# Plot reads by replicate and locus ------------------------------------
for (s in unique(rl_read_counts_serialdil2$specimen_id)) {
  title_lbl <- str_c("Sample", str_sub(s, start = 3L), sep = " ")
  rl_read_counts_serialdil2 %>%
    filter(specimen_id == s) %>%
    arrange(protocol, parasitemia) %>%
    mutate(
      sample_id = factor(sample_id, levels = unique(sample_id), ordered = TRUE)
    ) %>%
    read_count_heatmap(y_lbl = "Replicate", title_lbl = title_lbl) %>%
    print()
}
```

These heatmaps show the number of reads in the final dataset for each
locus-replicate combination. The plots have been grouped according to specimen.

```{r}
# Plot target length versus amplification ------------------------------
target_lengths <- locus_metadata %>%
  mutate(trg_length = end_pos - start_pos) %>%
  select(-locus_pos)
rl_read_counts_serialdil2 %>%
  group_by(locus) %>%
  summarize(median_n_read = median(n_read), .groups = "drop") %>%
  left_join(target_lengths, by = "locus") %>%
  ggplot(mapping = aes(x = trg_length, y = median_n_read)) +
  geom_point() +
  labs(x = "Target Length", y = "Median Read Count")
```

This plot shows the relationship between the length of the target and the median
read count.

# Filtering at Each Step

## UCI 12/23

```{r}
# Read in read filtering information -----------------------------------
reads_summary_uci1223 <- read_table(
    inp$reads_summary_uci1223, 
    col_names = c(
      "sample", 
      "input", 
      "filtered", 
      "denoisedF", 
      "denoisedR", 
      "merged", 
      "nonchim"
    ), 
    col_types = cols(.default = "i", sample = "c"), 
    skip = 1, 
    progress = FALSE
  ) %>%
  rename(sample_id = sample)

# Join final read counts -----------------------------------------------
reads_by_step_uci1223 <- rl_read_counts_uci1223 %>%
  group_by(sample_id) %>%
  summarize(n_read = sum(n_read), .groups = "drop") %>%
  right_join(reads_summary_uci1223, by = "sample_id") %>%
  rename(final = n_read) %>%
  relocate(final, .after = last_col()) %>%
  pivot_longer(-sample_id, names_to = "workflow_step", values_to = "n_read") %>%
  group_by(workflow_step) %>%
  summarize(n_read = sum(n_read), .groups = "drop") %>%
  mutate(
    workflow_step = factor(
      workflow_step, 
      c(
        "input", 
        "filtered", 
        "denoisedF", 
        "denoisedR", 
        "merged", 
        "nonchim", 
        "final"
      )
    )
  ) %>%
  arrange(workflow_step)
reads_by_step_uci1223
```

This table shows the total number of reads after each step in the DADA2 part of
the pipeline plus the final filtering done by `ASV_to_CIGAR.py`. At no point is 
there an order(s) of magnitude reduction in the data volume, which would 
indicate cause for concern.

## Serial Dilution Run 2

```{r}
# Read in read filtering information -----------------------------------
reads_summary_serialdil2 <- read_table(
    inp$reads_summary_serialdil2, 
    col_names = c(
      "sample", 
      "input", 
      "filtered", 
      "denoisedF", 
      "denoisedR", 
      "merged", 
      "nonchim"
    ), 
    col_types = cols(.default = "i", sample = "c"), 
    skip = 1, 
    progress = FALSE
  ) %>%
  rename(sample_id = sample)

# Join final read counts -----------------------------------------------
reads_by_step_serialdil2 <- rl_read_counts_serialdil2 %>%
  group_by(sample_id) %>%
  summarize(n_read = sum(n_read), .groups = "drop") %>%
  right_join(reads_summary_serialdil2, by = "sample_id") %>%
  rename(final = n_read) %>%
  relocate(final, .after = last_col()) %>%
  pivot_longer(-sample_id, names_to = "workflow_step", values_to = "n_read") %>%
  group_by(workflow_step) %>%
  summarize(n_read = sum(n_read), .groups = "drop") %>%
  mutate(
    workflow_step = factor(
      workflow_step, 
      c(
        "input", 
        "filtered", 
        "denoisedF", 
        "denoisedR", 
        "merged", 
        "nonchim", 
        "final"
      )
    )
  ) %>%
  arrange(workflow_step)
reads_by_step_serialdil2
```

This table shows the total number of reads after each step in the DADA2 part of
the pipeline plus the final filtering done by `ASV_to_CIGAR.py`. At no point is 
there an order(s) of magnitude reduction in the data volume, which would 
indicate cause for concern.

# Relationship with Parasitemia

## UCI 12/23

```{r}
# Compute and print sample read count and parasitemia ------------------
rl_pct_good_amp_uci1223 <- rl_read_counts_uci1223 %>%
  filter(DARC_Phenotype != "negative") %>%
  select(-DARC_Phenotype) %>%
  group_by(sample_id, specimen_id, lib_name, ct) %>%
  summarize(
    pct_loci_good_amp = (sum(n_read >= 10)/88) * 100,  
    .groups = "drop"
  ) %>%
  mutate(parasitemia = (10^((ct - 41.663)/-3.289)))

# Plot parasitemia versus sample read count ----------------------------
rl_pct_good_amp_uci1223 %>%
  group_by(specimen_id) %>%
  filter(pct_loci_good_amp == max(pct_loci_good_amp)) %>%
  ungroup() %>%
  pct_good_amp_vs_parasitemia()
```

This scatterplot shows the relationship between parasitemia and the percentage
of loci that have amplified successfully. The x-axis is $log_{10}$ scaled. The
line and shaded area show the linear relationship y = log(x) and the 95%
confidence interval for this model.

The two Ct replicates have been averaged, and samples from Duffy negative 
individuals have been removed. For each specimen, the replicate with the best
amplification was selected for inclusion in the plot.

## Serial Dilution Run 2

```{r}
# Compute and print sample read count and parasitemia ------------------
rl_pct_good_amp_serialdil2 <- rl_read_counts_serialdil2 %>%
  # Keep protocol "v2" over "v1" when both present
  filter(parasitemia != 100 | protocol == "v2") %>%
  select(-protocol) %>%
  group_by(sample_id, specimen_id, parasitemia) %>%
  summarize(
    pct_loci_good_amp = (sum(n_read >= 10)/88) * 100,  
    .groups = "drop"
  )

# Plot parasitemia versus sample read count ----------------------------
rl_pct_good_amp_serialdil2 %>%
  pct_good_amp_vs_parasitemia()
```

This scatterplot shows the relationship between parasitemia and the percentage
of loci that have amplified successfully. The x-axis is $log_{10}$ scaled. The
line and shaded area show the linear relationship y = log(x) and the 95%
confidence interval for this model.

For the 100 parasitemia preparations, those done with protocol v2 were selected
for visualization. Otherwise, all replicates for all samples are included in the
plot.

```{r}
# Save data for publication figures ------------------------------------
rl_read_counts_serialdil2 %>%
  write_csv(out$rl_read_counts)
rl_pct_good_amp_serialdil2 %>%
  write_csv(out$rl_pct_good_amp_parasitemia)
```
