---
title: Evaluation Power for Relatedness Estimation with paneljudge
author: Alfred Hubbard
output:
  html_notebook: default
  pdf_document: default
  github_document: default
---

```{r setup, echo = FALSE, message = FALSE, warning = FALSE}
# Load required libraries ----------------------------------------------
# These will be referenced without the `package::` construct, and thus 
# are loaded second to avoid masking
library(dplyr)
library(ggplot2)
library(knitr)
library(magrittr)
library(purrr)
library(readr)
library(stringr)
library(tidyr)

# Don't include code in PDF output
opts_chunk$set(echo = ! is_latex_output())
# Set default ggplot2 theme
theme_set(theme_bw())

# Hardcode parameters. These are not passed as shell arguments because 
# hardcoded values facilitate development and this document is not 
# meant to be usable with other projects.
inp <- list(
  rel_sim_res_dir = "../../results/rel_sim", 
  shared_functions = "../shared_functions.R"
)
out <- list(
  rel_sim_res_all = "../../results/rel_sim_res_all.tsv"
)

source(inp$shared_functions)
```

# Power for Relatedness Estimation

```{r out.width = "100%"}
# Read in simulated relatedness estimates ------------------------------
rel_sim_res <- tibble(
    sim_res_path = list.files(inp$rel_sim_res_dir, full.names = TRUE)
  ) %>%
  mutate(
    sim_res = map(
      sim_res_path, 
      read_tsv, 
      col_types = cols(
        .default = col_double(), 
        pair_num = col_integer()
      ), 
      progress = FALSE
    )
  ) %>%
  mutate(sim_res_path = basename(sim_res_path)) %>%
  mutate(sim_res_path = str_remove(sim_res_path, ".tsv")) %>%
  separate_wider_delim(sim_res_path, "_", names = c(NA, NA, "panel", "pop")) %>%
  unnest(sim_res)

# Add labels for plots -------------------------------------------------
pop_lbls <- tribble(
    ~pop, ~pop_lbl, 
    "AF", "Ethiopia", 
    "ESEA", "Cambodia and Vietnam", 
    "LAM", "Latin America"
  )
rel_sim_res <- rel_sim_res %>%
  left_join(pop_lbls, by = "pop") %>%
  mutate(panel_lbl = panel) %>%
  mutate(panel_lbl = str_replace("Kleinecke", "Kleinecke et al."))

# Compute and plot 95% CI widths of rhat by data-generating r ----------
rel_sim_res %>%
  rhat_ci_width_boxplot()
```

These boxplots show the distribution of CI widths for the relatedness estimate
for each simulated pair according to the data-generating r value. They are
facetted by region used for the allele frequencies. Different panels are shown
in different colors.

```{r out.width = "100%"}
# Compute and plot RMSEs for each data-generating r --------------------
rel_sim_res %>%
  plot_rhat_rmse()
```

These plots show the RMSE of rhat for each data-generating r value, facetted by
region. Different panels are shown in different colors.

```{r}
rel_sim_res %>%
  write_tsv(out$rel_sim_res_all)
```
