---
title: Evaluation of Marker Diversity with paneljudge
author: Alfred Hubbard
output:
  html_notebook: default
  pdf_document: default
  github_document: default
---

```{r setup, echo = FALSE, message = FALSE, warning = FALSE}
# Load required libraries ----------------------------------------------
library(dcifer)
library(paneljudge)
# These will be referenced without the `package::` construct, and thus 
# are loaded second to avoid masking
library(dplyr)
library(ggplot2)
library(knitr)
library(magrittr)
library(purrr)
library(readr)
library(stringr)
library(tidyr)

# Don't include code in PDF output
opts_chunk$set(echo = ! is_latex_output())
# Set default ggplot2 theme
theme_set(theme_bw())

# Hardcode parameters. These are not passed as shell arguments because 
# hardcoded values facilitate development and this document is not 
# meant to be usable with other projects.
inp <- list(
  mh_full = "../../results/microhap/MalariaGEN/mg_microhap.csv", 
  shared_functions = "../shared_functions.R"
)
out <- list(af_dir = "../../results/af")

source(inp$shared_functions)
```

# Marker Cardinality

```{r out.width = "100%", fig.height = 3}
# Calculate allele frequency with Dcifer
dcifer_calcafreq <- function(mh_tib) {

  # Convert allele frequency list to tibble for one target
  onetargetaf_list2tib <- function(onetargetaf) {
    tibble(seq = names(onetargetaf), freq = unname(onetargetaf))
  }

  # Compute allele frequency with Dcifer and reformat to tidy tibble
  # All samples should have a COI of 1
  coi <- rep(1, n_distinct(mh_tib$sample_id))
  af_dcifer <- mh_tib %>%
    dcifer::formatDat(svar = "sample_id", lvar = "locus", avar = "hapseq") %>%
    dcifer::calcAfreq(coi)
  tibble(
    target_id = names(af_dcifer), 
    alleles_freqs = unname(af_dcifer)
  ) %>%
  mutate(alleles_freqs = map(alleles_freqs, onetargetaf_list2tib)) %>%
  unnest(alleles_freqs)

}

# Calculate marker metrics given a tibble of allele frequencies
compute_marker_metrics <- function(af_long) {

  # Convert allele frequencies to matrix
  af_wide <- af_dcifer2wide(af_long)
  af_matrix <- af_wide %>%
    select(-target_id) %>%
    as.matrix()
  rownames(af_matrix) <- af_wide$target_id

  # Compute effective cardinalities
  # The warn_fs argument doesn't seem to work, so quietly() is used 
  # instead
  quiet_compute_eff_cardinalities <- quietly(
    paneljudge::compute_eff_cardinalities
  )
  eff_cardinalities_vec <- quiet_compute_eff_cardinalities(af_matrix)[[1]]
  eff_cardinalities_tib <- tibble(
      target_id = names(eff_cardinalities_vec), 
      eff_cardinality = unname(eff_cardinalities_vec)
    )
  af_long %>%
    group_by(target_id) %>%
    summarize(cardinality = n_distinct(seq), .groups = "drop") %>%
    left_join(eff_cardinalities_tib, by = "target_id")
}

# Read in microhaplotype data ------------------------------------------
mh_full <- read_csv(
    inp$mh_full, 
    col_types = cols(
      .default = col_character(), 
      Lat = col_double(), 
      Long = col_double(), 
      Year = col_integer(), 
      `% callable` = col_double(), 
      Fws = col_double(), 
      F_MISS = col_double()
    ), 
    progress = FALSE
  ) %>%
  filter(
    (Population == "ESEA") |
    (Population == "LAM") |
    (Country == "Ethiopia")
  ) %>%
  select(locus, sample_id, hapseq, Population) %>%
  mutate(panel = "PvGAP") %>%
  # Filter to diversity loci
  filter(str_detect(locus, "PvP01"))

# Compute allele frequency with Dcifer ----------------------------------
af <- mh_full %>%
  nest(mh_data = c(locus, sample_id, hapseq)) %>%
  mutate(allele_freqs = map(mh_data, dcifer_calcafreq)) %>%
  select(-mh_data) %>%
  unnest(allele_freqs)

# Write allele frequencies for use in relatedness simulations ----------
panels <- c("PvGAP")
pops <- unique(mh_full$Population)
for (panel in panels) {
  for (pop in pops) {
    out_path <- file.path(
      out$af_dir, 
      str_c("af", panel, str_c(pop, ".tsv"), sep = "_")
    )
    af %>%
      filter(panel == panel) %>%
      filter(Population == pop) %>%
      select(-panel, -Population) %>%
      write_tsv(out_path)
  }
}

# Compute marker metrics for each population ---------------------------
marker_metrics <- af %>%
  nest(af = c(target_id, seq, freq)) %>%
  mutate(marker_metrics = map(af, compute_marker_metrics)) %>%
  select(-af) %>%
  unnest(marker_metrics)

# Add population labels ------------------------------------------------
pop_lbls <- tribble(
    ~Population, ~pop_lbl, 
    "AF", "Ethiopia", 
    "ESEA", "Cambodia and Vietnam", 
    "LAM", "Latin America"
  )
marker_metrics <- marker_metrics %>%
  left_join(pop_lbls, by = "Population")

# Visualize cardinality versus effective cardinality -------------------
scale_max <- max(c(marker_metrics$cardinality, marker_metrics$eff_cardinality))
scale_lim <- c(0, scale_max)
marker_metrics %>%
  ggplot(mapping = aes(x = cardinality, y = eff_cardinality)) +
  geom_point(shape = 95, size = 3) +
  geom_abline(slope = 1, intercept = 0) +
  facet_grid(rows = vars(panel), cols = vars(pop_lbl)) +
  scale_x_continuous(limits = scale_lim) +
  scale_y_continuous(limits = scale_lim) +
  labs(x = "Cardinality", y = "Effective Cardinality")
```

This plot shows the cardinality, meaning the number of unique alleles at each
locus, versus the effective cardinality, which is a combined measure of the
cardinality and the allelic diversity. Effective cardinality cannot exceed
cardinality, so the y=x line indicates the maximum effective cardinality for
each locus.

This table shows the number of loci that have no variation in the entire
dataset:

```{r}
# Compute number of loci with no variation in entire dataset -----------
marker_metrics %>%
  group_by(panel, target_id) %>%
  summarize(n_pop_novar = sum(cardinality), .groups = "drop_last") %>%
  summarize(n_locus_novar = sum(n_pop_novar == 3), .groups = "drop")
```

This table shows the number of loci that have no variation in each population:
  
```{r}
# Compute number of loci with no variation by population ---------------
marker_metrics %>%
  group_by(panel, pop_lbl) %>%
  summarize(n_novar = sum(cardinality == 1), .groups = "drop")
```

This table shows the sum of all of the effective cardinalities, which is a 
measure of the overall informativeness of the panel in that population:

```{r}
# Compute "total cardinality" for each population ----------------------
marker_metrics %>%
  filter(cardinality > 1) %>%
  group_by(panel, pop_lbl) %>%
  summarize(total_cardinality = sum(eff_cardinality), .groups = "drop")
```

Loci with a cardinality of one were filtered out of each population prior to 
this calculation.
