---
title: Test for Selective Pressure on each Marker
author: Alfred Hubbard
output:
  html_notebook: default
  pdf_document: default
  github_document: default
---

```{r setup, echo = FALSE, message = FALSE, warning = FALSE}
# Load required libraries ----------------------------------------------
library(pegas)
# These will be referenced without the `package::` construct, and thus 
# are loaded second to avoid masking
library(knitr)
library(magrittr)
library(tidyverse)

# Don't include code in PDF output
opts_chunk$set(echo = ! is_latex_output(), out.width = "100%")
# Set default ggplot2 theme
theme_set(theme_bw())

# Hardcode parameters. These are not passed as shell arguments because 
# hardcoded values facilitate development and this document is not 
# meant to be usable with other projects.
inp <- list(
  mh_popbin = "../../results/microhap/MalariaGEN/mg_microhap_popbin.rds"
)
out <- list(trgs2filter = "../../results/trgs2filter_selec.csv")
```

The Tajima test was used to look for an impact of natural selection on the 
markers selected for this panel. Tajima exploited the fact that two metrics of 
genetic diversity, the number of segregating sites and the average number of 
pairwise nucleotide differences, differ in their response to selection to create 
a test for selection based on the difference between these two metrics.

Note that the malaria system violates the assumptions of the Tajima test, 
because mating is non-random and recombination is present. Tajima (1989) briefly 
discusses the impact of recombination on the test statistic, stating that the 
variance is reduced and the test therefore becomes more conservative. Our goal 
in testing for selection is to screen out loci that may not be neutral prior to 
performing downstream population genetic analyses. A too-conservative test for 
selection would potentially lead to data being removed unnecessarily. We are 
comfortable with this possibility, as we would rather filter out too much data 
than too little.

Given that the assumptions of the test are not met, we do not treat the results 
as concrete evidence that a given loci IS under selection - rather we consider 
it an indication that the loci MAY BE under selection. We interpret the results 
accordingly.

The test was performed with the pegas R package. It was performed separately for 
the data corresponding to each population, as defined by MalariaGEN.

Some of the tests generated warnings - the set of warning messages obtained is 
printed below:

```{r}
# Read in data ---------------------------------------------------------
mh_popbin <- read_rds(inp$mh_popbin)

# Perform Tajima test --------------------------------------------------
quiet_tajima <- quietly(pegas::tajima.test)
selec_res <- mh_popbin %>%
  mutate(selec_out = map(seqs_dnabin, quiet_tajima)) %>%
  mutate(selec_res = map(selec_out, pluck, "result")) %>%
  mutate(selec_warn = map(selec_out, pluck, "warnings")) %>%
  select(-seqs_dnabin)

# Analyze warnings -----------------------------------------------------
selec_res %>%
  select(target, selec_warn) %>%
  unnest(selec_warn) %$%
  unique(selec_warn)
```

This warning is expected for some of the target/population combinations, and is
not cause for concern.

The distribution of *p*-values obtained for these tests is given below.

```{r}
# Extract test statistic and p-value -----------------------------------
selec_res <- selec_res %>%
  mutate(taj_d = map_dbl(selec_res, pluck, "D")) %>%
  # According to Tajima (1989), the beta distribution typically 
  # provides a better approximation for the D statistic distribution 
  # than the normal distribution does
  mutate(pval = map_dbl(selec_res, pluck, "Pval.beta")) %>%
  select(-selec_out, -selec_warn, -selec_res) %>%
  # Remove tests with NaN for results, which means there were no 
  # segregating sites. These are excluded from the Bonferroni threshold 
  # calculation, with the logic that if data conditions prevent the 
  # test statistic being from being calculated, it doesn't really 
  # "count" as performing a hypothesis test.
  filter(! is.nan(taj_d))

# Visualize distribution of p-values -----------------------------------
selec_res %>%
  ggplot(mapping = aes(x = pval)) +
  geom_histogram(bins = 20) +
  labs(x = "p-value", y = "No. of Tests")
```

If the null hypothesis is true, a uniform distribution of *p*-values is
expected. This distribution is more normal than uniform, with a slight left
skew. This suggests that some of the tests may be significant, but likely not
many.

To assess significance of these results, the Bonferroni correction for multiple
testing is applied. This approach is sometimes considered too conservative, so a 
generous *p*-value of 0.05 was used to help compensate for that possibility. The 
Bonferroni-corrected significance threshold for this case is:

```{r}
# Apply Bonferroni correction ------------------------------------------
bf_thres <- 0.05 / nrow(selec_res)
bf_thres
selec_res_signif <- selec_res %>%
  filter(pval < bf_thres)
selec_res_signif %>%
  select(-chrom)
```

The table above shows the target/population combinations that yielded a 
significant result at this corrected threshold. These results suggest that 
pvcrt_o.10k.indel is under selective pressure in the WAS and ESEA populations, 
and PvP01_10_v1_1072001_1072200 is under selective pressure in the WAS 
population. These loci will be removed for downstream population genetic 
analyses that assume neutral markers.

```{r}
# Save list of targets to filter ---------------------------------------
selec_res_signif %>%
  select(target) %>%
  distinct() %>%
  write_csv(out$trgs2filter)
```
