---
title: "PvGAP: Development of a globally-applicable, highly-multiplexed 
microhaplotype amplicon panel and protocol for *Plasmodium vivax*"
subtitle: "Supplemental Figures"
author:
- Alfred Hubbard
- Edwin Solares
- Lauren Bradley
- Brook Jeang
- Delenasaw Yewhalaw
- Daniel Janies
- Eugenia Lo
- Guiyun Yan
- Elizabeth Hemming-Schroeder
output:
  pdf_document:
    number_sections: true
---

```{r setup, echo = FALSE, message = FALSE, warning = FALSE}
# Load required libraries ----------------------------------------------
# These will be referenced without the `package::` construct, and thus 
# are loaded second to avoid masking
library(knitr)

# Set default chunk parameters
opts_chunk$set(echo = FALSE, fig.align = "center", out.width = "100%")
```

```{r}
include_graphics("../../results/figs/suppfig1.pdf")
```

**Supplemental Figure 1:** Mean read counts across replicates for the serial dilutions,
visualized separately for each locus. Values in the color legend indicate break
points between bins. The $y$-axis tick labels indicate the specimen ID and
parasite copy number of each dilution, with the latter in units of parasite
copies/$\mu$L.


```{r out.width = "80%"}
include_graphics("../../results/figs/suppfig2.pdf")
```

**Supplemental Figure 2:** Within-sample allele frequencies of the genotypes observed from 
each replicate. Rows correspond to replicates for each specimen and parasite 
copy number combination (in units of parasite copies/$\mu$L). Columns correspond 
to loci. Colored rectangles show the within-sample allele frequency of each 
allele for that replicate and locus. Within each column, the same colors 
correspond to the same alleles for that locus. Across columns, the color 
palette is repeated.

```{r}
include_graphics("../../results/figs/suppfig3.pdf")
```

**Supplemental Figure 3:** Lin’s concordance correlation coefficient between replicate pairs 
for each parasite copy number. The bars around each point indicate the 95% 
confidence interval of the estimate. Different colors distinguish different 
samples. Note that not all samples were evaluated at all parasite copy 
numbers. The top facet shows the concordance for all alleles, whereas the 
bottom shows the concordance calculated only on alleles that are present in 
both replicates.

```{r}
include_graphics("../../results/figs/suppfig4.pdf")
```

**Supplemental Figure 4:** Distribution of the 95% confidence interval widths of $\hat{r}$ for each 
data-generating r value, as simulated by paneljudge. Facets separate results 
for different populations, and boxplot color indicates the marker panel. For 
clarity, only data-generating r values of 0.01, 0.15, 0.25, 0.5, and 0.99 
are included. The whiskers are Tukey-style and extend to a maximum of 1.5 * IQR.
